---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dtverse

<!-- badges: start -->
<!-- badges: end -->

Built on the top of data.table, `dtverse` is a grammar of data manipulation with data.table, providing a consistent a series of utility functions that help you solve the most common data manipulation challenges:

* Select columns
* Split one column to multiple columns based on patterns
* Filter cases based on their values
* Fill missing values
* Summarize and reduces multiple values down to a single summary
* Reshape `long to wide` or `wide to long`

## Installation

You can install from github:

``` r
install.packages("dtverse")
```

## Select columns
Select variables in a data table. You can also use predicate functions like is.numeric to select variables based on their properties (e.g. 1:3 selects the first column to the third column).

```{r ex1}
library(dtverse)
library(data.table)
data("dt_dates")
dt_dates <- setDT(dt_dates)
dtverse::select_cols(dt_dates, c("Start_Date", "Full_name"))
```


## Split a column
Split a column with its special pattern, and assign to multiple columns respectively. For example, split full name column to first name and last name column. 

```{r ex2}
data("dt_dates")
library(data.table)
data("dt_dates")
dtverse::str_split_col(dt_dates,
              by_col = "Full_name",
              by_pattern = ", ",
              match_to_names = c("First Name", "Last Name"))
```


## Filter cases based on values

`filter_all()` is to return a data table with **ALL** columns (greater than/ less than/ equal to) a desired value. 

```{r ex3}
data("dt_values")
dtverse::filter_all(dt_values, operator = "l", .2)
```

`filter_any()` is to return a data table with **ANY** columns (greater than/ less than/ equal to) a desired value. 


```{r ex4}
data("dt_values")
dtverse::filter_any(dt_values, operator = "l", .1)
```

Similarly, `filter_all_at()` is to return a data table with **ALL selected** columns (greater than/ less than/ equal to) a desired value. 

```{r ex5}
data("dt_values")
dtverse::filter_all_at(dt_values, operator = "l", .1, c("A1", "A2"))
```

Similarly, `filter_any_at()` is to return a data table with **ANY selected** columns (greater than/ less than/ equal to) a desired value. 


```{r ex6}
data("dt_values")
dtverse::filter_any_at(dt_values, operator = "l", .1, c("A1", "A2"))
```

## Fill missing values

`fill_NA_with()` will fill NA value with a desired value in the selected columns. If `fill_cols` is `All` (same columns type), it will apply to the whole data table.

```{r ex7}
data("dt_missing")
dtverse::fill_NA_with(dt_missing, fill_cols = c("Full_name"), fill_value = "pending")
```

## Group by and summarize

`dt_group_by()` is to group by desired columns and summarize rows within groups. 


```{r ex8}
data("dt_groups")
print(head(dt_groups))
```

Now we see the `dt_groups` data table has A1, A2 as numeric columns, and group1, group2 as group infomation. 

```{r ex9}
data("dt_groups")
dtverse::dt_group_by(dt_groups, 
            group_by_cols = c("group1", "group2"), 
            summarize_at = "A1", 
            operation = "mean")
```

Now we want to group by group1 and group2, then fetch the first within each group, we can use `get_row_group_by()` function. 

```{r ex10}
data("dt_groups")
dtverse::get_row_group_by(dt_groups, 
                 group_by_cols = c("group1", "group2"), 
                 fetch_row = "first")
```

or last row with same example. 

```{r ex11}
data("dt_groups")
dtverse::get_row_group_by(dt_groups, 
                 group_by_cols = c("group1", "group2"), 
                 fetch_row = "last")
```


## Reshape `long to wide` or `wide to long`

Here is an example of reshaping a data table from wide to long. 

```{r ex12}
data("dt_dates")
print(head(dt_dates))
dtverse::reshape_longer(dt_dates, 
               keep_cols = "Full_name", 
               by_pattern = "Date", 
               label_cols = c("Date_Type"), 
               value_cols = "Exact_date", 
               fill_NA_with = NULL)
```

Here is an example of reshaping a data table from long to wide. 

```{r ex13}
data("dt_long")
print(head(dt_long))
dtverse::reshape_wider(dt_long, 
              keep_cols = c("Full_name"), 
              col_lable = c("Date_Type"), 
              col_value = "Exact_date")
```

